\section{Hello World - an introduction to the syntax}
CQL is strictly a language designed to implement conjunctive queries on local data, so a 'Hello World!' program may appear slightly odd, however in this section the syntax of the language will be introduced. 


\subsection{Writing your first program}
At it's core, CQL allows the selection of data in CSV files. In this example program, we'll be loading data from two CSV files, P.csv and Q.csv, and outputting all rows in the file where the first value of each row is equal. Given two CSV files:


\setbox0=\hbox{%
\begin{minipage}{1.9in}
\lstinputlisting[
basicstyle={\ttfamily},
identifierstyle={\color{black}},
tabsize=2,
numbersep=8pt,
numbers=left,
title=P.csv,
xleftmargin=0.5cm,
frame=tlbr,
framesep=2pt,
framerule=0pt,
morekeywords ={class,run}
]{P.csv}
\end{minipage}
}
\savestack{\listingA}{\box0}

\setbox0=\hbox{%
\begin{minipage}{1.9in}
\lstinputlisting[
basicstyle={\ttfamily},
identifierstyle={\color{black}},
tabsize=2,
numbersep=8pt,
numbers=left,
title=Q.csv,
xleftmargin=0.5cm,frame=tlbr,framesep=2pt,framerule=0pt,
morekeywords ={class,run}
]{Q.csv}
\end{minipage}
}
\savestack{\listingB}{\box0}

\begin{center}
\begin{tabular}{|c|c|}
\hline
\stackinset{c}{5pt}{t}{}{}{\listingA} &
\stackinset{c}{5pt}{t}{}{}{\listingB} \\
\hline
\end{tabular}
\end{center}
\begin{normalsize}
We can use the equality operator \codeword{=} to output only rows where P.csv and Q.csv have the same value. We will introduce the full syntax for conjunctive queries later, but formally this is written as:
$$
	x_1, x_2 \vdash P(x_1) \land Q(x_2) \land x_1 = x_2 
$$
In order to implement this in CQL, we need a program as below:
\lstinputlisting[style=framed]{HelloWorld.cql}
Now that we've got an extremely simple first program, we can begin to run programs using our interpreter.
\end{normalsize}
\subsection{Running your first program}
\begin{normalsize}
CQL programs are saved with the file extension \textbf{.cql}, and can be run by placing the executable interpreter in the same directory as your data (in the form of .CSV files), and executing your program as:
\begin{lstlisting}
	$./CQL HelloWorld.cql
\end{lstlisting}
Where 'CQL' is the name of your executable, and 'HelloWorld.cql' the name of your program. Note that on Windows systems, running the executable will have the syntax:
\begin{lstlisting}
	CQL.exe HelloWorld.cql
\end{lstlisting}
If all goes well, there should be an output identical to the one below; if your program throws errors, some common troubleshooting steps are ensuring that the CSV files, CQL programs and executable are all in the same directory, and that your executable has executable permission on Unix based systems. There are also some descriptions of common errors in \textbf{Sections 3 and 4}.
\end{normalsize}
\begin{lstlisting}[style=framed]
1,1
2,2
2,2
\end{lstlisting}
\begin{normalsize}
Congratulations! You've executed your first CQL program. 
\end{normalsize}